\subsection{A naive, retrospective, optimal choice}

We have observed that the location an \event takes place in significantly
impacts the distribution of origin of its participants. However, dismissing
temporarily this factor to consider what could have been the cheapest location
for past conferences can be an illuminating exercise.

To this end, we chose a fixed number of locations that we believe to be
representative and spread across the relevant parts of the globe for our
concern: Paris, Edinburgh, Boston, Los Angeles, Vancouver, Tokyo, Beijing and
Mumbai. We then reprocessed the dataset to look for the location that would have
led to the lowest carbon footprint, once again assuming that it would not have
changed the set of participants.

\begin{figure}
  \begin{tabular}{|l|l|c|c|c|c|}
    \hline%
    \bfseries event & \bfseries orig. loc. & \bfseries orig. cost & \bfseries best loc. & \bfseries best cost & \bfseries saved
    \csvreader[head to column names]{../output/sigplan/optimals.csv}{}%
              {\\\conf\ \year & \csvcoliii & \csvcoliv & \csvcolv & \csvcolvi & \csvcolvii}%
              \\\hline
  \end{tabular}
  \caption{For each \event, depicts the location among Paris, Edinburgh,
    Boston, Los Angeles, Vancouver, Tokyo, Beijing and Mumbai that would
    have led to the lowest carbon footprint. We indicate 'SAME' when the
    actually location is the best, and show in the final column the amount
    of \gazunitbis that it would have saved. \bcp{Might be more informative
      to replace SAME with the name of the city, in parens or something to
      indicate the sameness.}}
  \label{fig:optimal}
\end{figure}


Figure~\ref{fig:optimal} depicts the resulting data: for each \event, the best location
and the average \gazunitbis it would have saved. We observe that in the majority
of the events, the locality effect is strong enough that the optimal takes place on
the original continent. However it is striking to see how often the East Coast turns
out to be the cheapest destination. In particular it appears to be preferable to
the West Coast in most cases, even despite the underlying locality effect that
is ignored here.

